---
title: "Modular design and software development best practices in Python"
institute: "Odum Institute<br/>University of North Carolina at Chapel Hill"
format:
    revealjs:
        theme: [default, unc.scss]
        width: 1920
        height: 1080
        logo: UNC_logo_RGB.png
        slide-number: true
        footer: "https://projects.indicatrix.org/odum-modular-design-python"
execute:
    include: true
    eval: true
    echo: true
    output: true
jupyter: "air-analysis"
---

```{python}
#| include: false
#| echo: false
#| output: false

import pandas as pd
import numpy as np

data = pd.read_csv("../data/air_sample.csv")

market_ids = pd.read_csv("../data/L_CITY_MARKET_ID.csv")

data = data.merge(
    market_ids.rename(columns={"Description": "OriginCity"}).set_index("Code"),
    left_on="OriginCityMarketID",
    right_index=True
)

data = data.merge(
    market_ids.rename(columns={"Description": "DestCity"}).set_index("Code"),
    left_on="DestCityMarketID",
    right_index=True
)

carriers = pd.read_csv("../data/L_CARRIERS.csv")

data = data.merge(
    carriers.rename(columns={"Description": "OperatingCarrierName"}).set_index("Code"),
    left_on="OpCarrier",
    right_index=True
)

data = data.merge(
    carriers.rename(columns={"Description": "TicketingCarrierName"}).set_index("Code"),
    left_on="TkCarrier",
    right_index=True
)
```

## About me

- Assistant Professor of City and Regional Planning
    - Research focus: transportation modeling and simulation
- Quantitative Methods Consultant at Odum
- Three years as a software developer before graduate school

## Why this matters

- Code is increasingly common in research
- Most academics receive little to no formal training in software development
- This can lead to
    - Inefficient development (takes too much time)
    - Errors in code
    - Non-reproducible analyses

## What we'll be doing today

- Hands-on exercise converting a set of R scripts a modular and tested format
- Two scripts that analyze data on airline tickets from the Bureau of Transportation Statistics
- Script download: [https://github.com/mattwigway/odum-modular-design-python](https://github.com/mattwigway/odum-modular-design-python)
- These slides: [https://projects.indicatrix.org/odum-modular-design-python](https://projects.indicatrix.org/odum-modular-design-python)

## The two notebooks

- Both load and clean the air ticket data
- One analyzes the busiest air routes in the country
- One looks at airline "fortress hubs"---which airports have the most concentrated service from a single airline

## What we'll address

- Duplicated code
- Literal numeric values
- Automated tests
- Package version incompatibilities

## What we won't address

- Version control
    - Odum offers separate Git and Github class

## `Air analysis.ipynb`

- Hands-on walkthrough of file

## `Airline analysis.ipynb`

- Hands-on walkthrough of file

## Duplicated code (don't repeat yourself)

- Common software development rule: don't repeat yourself (DRY)
- In `Air analysis.ipynb` we do the same analysis twice, once by airport and once by city
- We can eliminate this duplicated code using a _function_

## Functions

- A function is a reusable piece of code that can be executed many times
- Functions separate the _definition_ of code from the _execution_
- You use functions all the time
    - For instance, `sum()` is a function

## Anatomy of a function

- Functions are Python code that is given a name to be referred to later
- Functions have three main components
    - Body
        - The main Python code of the function
    - Arguments
        - Values that can be specified each time the function is used
    - Return value
        - The result of the function

## Anatomy of a function

```{python}
import math

def calculate_cylinder_volume (radius, height):
    base_area = math.pi * radius ** 2
    volume = base_area * height
    return volume
```

::: {.incremental}
- _radius_ and _height_ are arguments
- Next three lines are body (the actual calculations)
- _volume_ is the return value
    - `return` ends the function, any code after `return` will not be run
- Python uses _indentation_ to define _blocks_—the body of the function is an _indented block_
    - Similarly, the body of an `if`, `for`, `with`, `while` statement is an indented block
:::

## Using functions

- Code above _defines_ the function
- It doesn't run any of the function code
- Function code is run when the function is _called_

. . . 

```{python}
calculate_cylinder_volume(1, 2)
```

## Defining functions in `Air analysis.ipynb`

- The analysis is repeated twice, once by airport pairs and once by city pairs
- We will re-write this analysis into a function, and use the same code twice

## Defining functions in `Air analysis.ipynb`

```{python}
#| eval: false
#| output: false
data["airport1"] = np.where(data.Origin < data.Dest, data.Origin, data.Dest)
data["airport2"] = np.where(data.Origin < data.Dest, data.Dest, data.Origin)

pairs = (
    data
        .groupby(["airport1", "airport2"], as_index=False)
        .agg({"Passengers": "sum", "Distance": "first"})
        .sort_values("Passengers", ascending=False)
)
```

## Step one: make it a function

- Move the relevant code to a single cell
- Indent it
- Prefix it with `def busiest_routes ():`
- Add `return pairs` at the end

```{python}
#| output: false
def busiest_routes ():
    data["airport1"] = np.where(data.Origin < data.Dest, data.Origin, data.Dest)
    data["airport2"] = np.where(data.Origin < data.Dest, data.Dest, data.Origin)

    pairs = (
        data
            .groupby(["airport1", "airport2"], as_index=False)
            .agg({"Passengers": "sum", "Distance": "first"})
            .sort_values("Passengers", ascending=False)
    )

    pairs["distance_km"] = pairs.Distance * 1.609

    return pairs
```

## Running our function

- Running the code on the previous slide defines the function, but does not run it
- We have to _call_ the function to run it

```{python}
busiest_routes()
```

## Adding arguments

- We want to make the `Origin` and `Dest` column names arguments to the function
- It's also standard to make the data itself and anything else you refer to inside the function arguments

## Adding arguments

```{python}
#| output: false
def busiest_routes (dataframe, origcol, destcol):
    dataframe["airport1"] = np.where(dataframe.origcol < dataframe.destcol, dataframe.origcol, dataframe.destcol)
    dataframe["airport2"] = np.where(dataframe.origcol < dataframe.destcol, dataframe.destcol, dataframe.origcol)

    pairs = (
        dataframe
            .groupby(["airport1", "airport2"], as_index=False)
            .agg({"Passengers": "sum", "Distance": "first"})
            .sort_values("Passengers", ascending=False)
    )

    pairs["distance_km"] = pairs.Distance * 1.609

    return pairs
```

## Adding arguments

```{python}
#| error: true
busiest_routes(data, "Origin", "Dest")
```

## What happened?

- Python tried to group by the columns `origcol` and `destcol`
- These columns don't exist
- We need to tell Python that `origcol` and `destcol` aren't actually column names, but variables that contain column names
- We can use subscript syntax (`["columnname"]`) to solve this problem
- Note that with subscript syntax, the column name is in quotes - this is a _literal string_
- Any place where we're using quotes, we can replace it with a variable name and the _content_ of that variable will be used

## Adding arguments

```{python}
#| output: false
def busiest_routes (dataframe, origcol, destcol):
    dataframe["airport1"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[origcol], dataframe[destcol])
    dataframe["airport2"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[destcol], dataframe[origcol])

    pairs = (
        dataframe
            .groupby(["airport1", "airport2"], as_index=False)
            .agg({"Passengers": "sum", "Distance": "first"})
            .sort_values("Passengers", ascending=False)
    )

    pairs["distance_km"] = pairs.Distance * 1.609

    return pairs
```

## Adding arguments

```{python}
busiest_routes(data, "Origin", "Dest")
```

## Using our function for cities

```{python}
busiest_routes(data, "OriginCity", "DestCity")
```

## Putting it all together

```{python}
#| output: false
def busiest_routes (dataframe, origcol, destcol):
    dataframe["airport1"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[origcol], dataframe[destcol])
    dataframe["airport2"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[destcol], dataframe[origcol])

    pairs = (
        dataframe
            .groupby(["airport1", "airport2"], as_index=False)
            .agg({"Passengers": "sum", "Distance": "first"})
            .sort_values("Passengers", ascending=False)
    )

    pairs["distance_km"] = pairs.Distance * 1.609

    return pairs

busiest_routes(data, "Origin", "Dest")
```

## Mutating arguments

- Our `busiest_routes` function _mutates_ its argument `dataframe` - i.e. it changes it
- We add `airport1` and `airport2` columns - and you can check, you can see these columns in your data
- This is generally a bad practice - functions should not change their arguments unless they are _explicitly_ designed to do so
- What if you were using a column called `airport1` already for something else?

## Mutating arguments

- Whatever happens inside the function should be isolated, changes to how the function works internally should not change anything about what the caller sees
- This already works for variables - if you delete your `pairs` variable from before we had a function (`{python eval=false} del pairs`), and run your `busiest_routes` function, you will not get a new `pairs` variable

## Mutating arguments

- The simplest fix is to add `dataframe = dataframe.copy()` at the top of your function
- If your data is not huge, this will work fine

## Mutating arguments

```{python}
def busiest_routes (dataframe, origcol, destcol):
    dataframe = dataframe.copy()
    dataframe["airport1"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[origcol], dataframe[destcol])
    dataframe["airport2"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[destcol], dataframe[origcol])

    pairs = (
        dataframe
            .groupby(["airport1", "airport2"], as_index=False)
            .agg({"Passengers": "sum", "Distance": "first"})
            .sort_values("Passengers", ascending=False)
    )

    pairs["distance_km"] = pairs.Distance * 1.609

    return pairs

busiest_routes(data, "Origin", "Dest")
```

## Default arguments

- Sometimes, you will have a function with a lot of arguments, but most of them don't need to be changed often
    - e.g. parameter values in a model estimation
- You specify these by just adding the default value with an `=` when defining the function
- When the function is called, the default value will be used unless a different value was specified

## Default arguments

```{python}
def busiest_routes (dataframe, origcol="Origin", destcol="Dest"):
    dataframe = dataframe.copy()
    dataframe["airport1"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[origcol], dataframe[destcol])
    dataframe["airport2"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[destcol], dataframe[origcol])

    pairs = (
        dataframe
            .groupby(["airport1", "airport2"], as_index=False)
            .agg({"Passengers": "sum", "Distance": "first"})
            .sort_values("Passengers", ascending=False)
    )

    pairs["distance_km"] = pairs.Distance * 1.609

    return pairs

busiest_routes(data)
```

## Default arguments

- But you can still specify the argument values

```{python}
busiest_routes(data, "OriginCity", "DestCity")
```

## Named arguments

- When calling a function, you can specify the arguments by name instead of by putting them in order
- When there are a lot of arguments, this can be very helpful
- You don't need to do anything special to your function to enable this

```{python}
busiest_routes(data, destcol="OriginCity", origcol="DestCity")
```

## Exercise: do the same in `Airline analysis.ipynb`

- Write a function to compute market shares
- Use that function to compute market shares by operating carrier and ticketing carrier
- Bonus: also compute market shares for individual airports instead of cities

## Result

```{python}
def market_shares(dataframe, carriercol, citycol):
    mkt_shares = (
        dataframe
            .groupby([citycol, carriercol])
            .Passengers
            .sum()
            .reset_index()
    )
    
    mkt_shares["market_share"] = mkt_shares.Passengers / mkt_shares.groupby(citycol).Passengers.transform("sum")
    
    mkt_shares = mkt_shares.sort_values("market_share", ascending=False)
    
    return mkt_shares.loc[mkt_shares.Passengers > 1000]
```

. . .

- Do we need to have a `dataframe = dataframe.copy()` line here?

## Result

```{r}
market_shares(data, "OperatingCarrierName", "OriginCity")
```

## Result

```{r}
market_shares(data, "TicketingCarrierName", "OriginCity")
```

## Result

```{r}
market_shares(data, "TicketingCarrierName", "Origin")
```


## Symbolic constants

- We still have the value `1.609` in our function to convert miles to kilometers
- It's a bad practice to have literal values like these in your code
    - In a large project, value will end up spread over many files
    - Difficult to change (maybe you want to increase accuracy later by changing it to `1.609344`)
    - Numbers used in many places can be prone to typos

## Symbolic constants

- We can define a _constant variable_ in our file that contains the value `1.609`, and refer to it when needed
- By convention, constants are all uppercase, and usually appear near the start of the file
- Python does not enforce variables being constant, you could change them in other code, but you shouldn't

## Symbolic constants

- Near the top of your notebook, add

```{r}
KILOMETERS_PER_MILE = 1.609
```

- Replace `1.609` in your function with `KILOMETERS_PER_MILE`

## Runtime checks: `assert`

- A first step towards ensuring correctness is to add checks throughout your code
- In Python, this is most easily done with the `assert` statement - this will error if whatever condition it checks is not true

## Runtime checks: `assert`

- Let's make sure all of the passenger numbers are positive and non-NA
- Add these lines to the start of your `busiest_routes` function:

```{python}
#| eval: false
assert (dataframe.Passengers >= 1).all()
assert not pd.isnull(dataframe.Passengers).any()
```

## Runtime checks: `assert`

```{python}
#| output: false
def busiest_routes (dataframe, origcol, destcol):
    # check assumptions
    assert (dataframe.Passengers >= 1).all()
    assert not pd.isnull(dataframe.Passengers).any()

    dataframe["airport1"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[origcol], dataframe[destcol])
    dataframe["airport2"] = np.where(dataframe[origcol] < dataframe[destcol], dataframe[destcol], dataframe[origcol])

    pairs = (
        dataframe
            .groupby(["airport1", "airport2"], as_index=False)
            .agg({"Passengers": "sum", "Distance": "first"})
            .sort_values("Passengers", ascending=False)
    )

    return pairs

busiest_routes(data, "Origin", "Dest")
busiest_routes(data, "OriginCity", "DestCity")
```

## Confirming that errors occur

```{python}
#| error: true
error_data = data.copy()
data.iloc[0, "Passengers"] = -1
busiest_routes(mutate(data, Passengers=-1), Origin, Dest)
```

## Using functions in multiple files

- Often we may want to use functions in multiple files
- For instance, we use the same data loading code in both notebooks
- The simplest way is to put the functions in another file, and use `import` to load them
- `import` will run the code in a file and make all of the functions and variables available
- Just as if that code was pasted into each source file

## Creating a data cleaning function

- Create a function that takes arguments for the file names of the dataset and the two tables that are joined to it,
    - performs the data cleaning/processing,
    - and returns the final dataset
- Save this function to a new file in the same directory as the others
- This should be a Python (`.py`) file, not a notebook
- You will need to import any functions that you use from other libraries (e.g. `pd.read_csv`)

## Using your data cleaning function

- Delete the data loading code in each of the analysis files
- Restart the kernel (Kernel -> Restart Kernel) to remove existing data from memory
- Add `import load` at the top of the file to import your new file
- Call your new function like this

```{python}
#| eval: false
data = load.load_data("data/air_sample.csv", "data/L_CITY_MARKET_ID.csv", "data/L_CARRIERS.csv")
```

## The finished function

Yours might not look exactly the same, and that's okay

```{python}
#| eval: false
import pandas as pd

def load_data(main_file, market_file, carrier_file):
    data = pd.read_csv(main_file)
    
    market_ids = pd.read_csv(market_file)

    data = data.merge(
        market_ids.rename(columns={"Description": "OriginCity"}).set_index("Code"),
        left_on="OriginCityMarketID",
        right_index=True
    )
    
    data = data.merge(
        market_ids.rename(columns={"Description": "DestCity"}).set_index("Code"),
        left_on="DestCityMarketID",
        right_index=True
    )

    carriers = pd.read_csv(carrier_file)
    
    data = data.merge(
        carriers.rename(columns={"Description": "OperatingCarrierName"}).set_index("Code"),
        left_on="OpCarrier",
        right_index=True
    )
    
    data = data.merge(
        carriers.rename(columns={"Description": "TicketingCarrierName"}).set_index("Code"),
        left_on="TkCarrier",
        right_index=True
    )

    return data
```

## All data manipulation should be in code

- I _never_ modify original data files, and rarely save new ones
- I always write code or functions to make all the changes/cleaning I need
- I run that function to load the data each time
- This way I can always see what has been done to data

## Creating Python packages

- Technically, what you did when you created `load.py` was create a package
- Just like `pandas` or `numpy`
- Pretty soon, though, you're not going to want to have all of your package code in one file

## Differences between Python scripts/notebooks and package code

- Almost everything in a package should be inside a function
    - Code that is not in functions will be run when the package is loaded
    - Main exception is constants

## Creating multi-file Python packages

- You use packages all the time when you use Python
    - Any time you run `import ...` you're loading a package
- You can also create packages yourself

## Advantages of creating a package

- Easy to share code, even between projects (no copying .py files to load with `import`)
- Packages facilitate automated testing

## Creating a multi-file package

- In its simplest form, a multi-file Python package is just a folder containing a file `__init__.py`
    - Note: double underscores before and after
- The folder name should be the name of the package

## Creating a multi-file package

- When you run `import package`, the code in `__init__.py` is run
- Any functions or variables defined there will be available

## Creating a multi-file package

- The whole point was to have multiple files, so what good is `__init__.py`?
- You can add functions in other files, and then import them in `__init__.py`

## Creating a multi-file package

- Call your package `airlinestats`
- Move your `load.py` file into your package folder
- Create a new file `analysis.py` and another `constants.py`
- Each of these files is its own package, often called a "module"

## Sidebar: namespaces

- Multiple packages may define functions with the same name
- To keep this straight, each Python package/module has its own _namespace_; names defined in one file are not accessible in another file
    - Functions also have namespaces; names used inside a function are not visible outside the function
- Functions and values must be imported to be used in a different file

## Sidebar: namespaces

- The most common `import` statement is just `import package`
    - or `import package as pkg`
- This requires you to type `package.function` every time you want to use something from that package's namespace
- You can also run `from package import function` to bring something directly into your package's namespace
    - `from package import *` to import everything from a package, but this can lead to unexpected results
- Then you can just type `function` to refer to it

## Adding Python code to your `airlinestats` package

- Put the functions to calculate market shares and route popularity in `analysis.py`
- Put the symbolic constant `KILOMETERS_PER_MILE` in `constants.py`
    - Good practice to put your constants in one file so they don't get defined in multiple places accidentally

## Loading your package

- You can import your package, just like you did before
- Run `import airlinestats` in your notebook

## Using your package

- Try running the `load_data` function

```{r}
#| eval: false
data = airlinestats.load_data(
    "data/air_sample.csv",
    "data/L_CITY_MARKET_ID.csv",
    "data/L_CARRIERS.csv"
)
```

. . . 

<pre>AttributeError: module 'airlinestats' has no attribute 'load_data'</pre>

## Making functions from other files available

- Remember that Python only runs the code in `__init__.py` when you import your package
- In `__init__.py`, we need to tell Python to make the `load_data`, `busiest_routes`, and `market_shares` functions available

## Making functions from other files available

```{python}
#| eval: false
from .load import load_data
from .analysis import *
```

- The `.` means the `.load` package _inside the current package_

## Try it again

- To reload the package, you need to either
    - Restart your kernel
    - Run `importlib.reload(airlinestats)`

## Try it again

- `load_data` should work now
- Now, try `busiest_routes` or `market_shares`

## Try it again

```{python}
#| eval: false
airlinestats.busiest_routes(data, "Origin", "Dest")
```

. . . 

<pre>NameError: name 'np' is not defined</pre>

## Fixing the issue

- We need to explicitly import numpy into the `analysis.py` file where we use it
- Add to the top of `analysis.py`:
    - `import numpy as np`
- Did that solve the problem?
- You may need to restart the kernel

. . .

<pre>NameError: name 'KILOMETERS_PER_MILE' is not defined</pre>

## Fixing the issue

::: {.incremental}
- We need to import the `KILOMETERS_PER_MILE` constant into the `analysis.py` file
- `from .constants import KILOMETERS_PER_MILE`
:::

## Installing the package

- To install our packages so that other projects on the system can access them, we need to make a slight change to our package
- First, we need to create another directory with the name of the package, enclosing the current `airlinestats` directory
- So it will look something like this:
- `airlinestats`
    - `airlinestats`
        - `__init__.py`
        - `analysis.py`
        - `constants.py`
        - `load.py`

## Installing the package

- We also need to give Python some information about the package itself before it can be installed
- We do this by creating a `pyproject.toml` file
- This file can contain lots of information (dependencies, build system configuration, etc.), but in this case we will just put in the package name and version

```
[project]
name = "airlinestats"
version = "0.0.1"
```

- Save this in the outer `airlinestats` directory

## Installing the package

- In a console, in the outer `airlinestats` directory, run `python -m pip install .`
- `airlinestats` is now installed for all project (or at least, all that use the current conda environment - more on that soon)

## Bringing it full circle

- Return to the original notebook files
- Rewrite them to use your library

## Advanced topics in packaging

- When you install a Python package with `conda` or `pip`, it downloads a lot of other packages that are needed as well—the _dependencies_
- We'll talk about project-level dependency management when we talk about `conda` environments, but if you were publishing a package you'd want to put dependency information in `pyproject.toml`

## Automated testing

- Automated testing formalizes testing and makes it repeatable
- You write code that tests your functions, and ensures they give the correct results
- Every time you change the code, you can run all the tests you've ever written, to make sure you haven't reintroduced old issues

## The structure of testing

- Automated tests consist of code and _expectations_ or _assertions_
- A test exercises one of the functions in your package, and specifies expectations about its output
- Good tests are small, with one piece of code tested and a handful of expections
- This may require splitting your package code up into multiple functions
- It's a good idea to test how your package acts when used incorrectly as well—for instance, when fed bad data

## Automated testing in Python

- There are many testing libraries for Python
- We'll use the built-in [unittest module](https://docs.python.org/3/library/unittest.html)

## Automated testing in Python

- We now have a `tests/` directory, with a `testthat` directory inside it
- Any file starting with `test` and ending with `.R` in this directory will be run as part of automated testing

## Writing tests

- Create an R file to hold tests for the analysis functions
- It's good practice to match the names of the test files to the names of the files that define the code being tested, so create `tests/testthat/test-analysis.R`

## Writing tests

- We'll write two tests for the analysis code
    1. We'll test that market shares of all airline at an airport sum to 1
    2. We'll test that passing in a dataset without a `Passengers` column leads to an error

## Writing tests

- Often, we need data for our tests
- Here, we will create a small dataset within our test file directly
- It is also possible to distribute a CSV with your package with test data
- I like to put my code to generate test data in a function, and re-create the data for each test, so that there is no chance of one test modifying the data used by another
    - Unless the data take so long to create or load that this is impractical
- Create a folder `test` inside our main top-level `airlinestats` folder

## Write the test for market shares

- Python unit tests use test _classes_ ("objects" that have functions that can be used with them)
- Every test class has to _inherit_ from the `unittest.TestCase` class
- Any function within the test class that starts with `test_` is a test case
- There are many `assertions` you can make about the results
- `setUp()` is a function run before each test case to generate data, etc.
- Let's create a file `test.py`
    - You don't have to call it this, and [most large projects will have tests spread across many files](https://github.com/mattwigway/eqsormo/tree/main/tests)
- Also create an empty `__init__.py` file in the `test` directory

## Write the tests

```{python}
#| eval: false
import unittest
import pandas as pd
import numpy as np
import airlinestats

class AirlineStatsTest(unittest.TestCase):
    def setUp(self):
        self.data = pd.DataFrame({
            "Origin": ["SFO", "ORD", "DCA", "PHX", "BOS", "RDU"] * 20,
            "Destination": ["DFW", "MCI", "MIA", "ABQ", "CLT"] * 24,
            "Passengers": [200, 70, 300, 300, 500] * 24,
            "Carrier": ["United", "American", "Delta", "Southwest", "JetBlue"] * 24
        })

    def test_market_shares(self):
        result = airlinestats.market_shares(self.data, "Carrier", "Origin")
        # now we group by airport to make sure the market shares are one
        overall_share = (
            result
                .groupby("Origin")
                .market_share
                .sum()
        )

        self.assertEqual(len(overall_share), 6)
        self.assertTrue(np.allclose(overall_share, 1))

    def test_passenger_column_missing(self):
        data = self.data.drop(columns=["Passengers"])
        with self.assertRaises(AttributeError):
            airlinestats.market_shares(data, "Carrier", "Origin")
```

## Run tests

- In a terminal, navigate to the top-level `airlinestats` directory
- Run `python -m unittest`

```
======================================================================
FAIL: test_market_shares (test.test_data.AirlineStatsTest.test_market_shares)
----------------------------------------------------------------------
Traceback (most recent call last):
  File "/Users/mwbc/git/odum-modular-design-python/airlinestats/test/test_data.py", line 29, in test_market_shares
    self.assertTrue(np.allclose(overall_share, 1))
AssertionError: False is not true

----------------------------------------------------------------------
Ran 2 tests in 0.006s

FAILED (failures=1)
```

## Fix test errors

- We got a test failure for the market shares test
- This is because we filtered out small airlines in the function
- Let's remove the filtering from the function
    - It's good to break your functions into smaller pieces for easier testing
- Re-run the tests

## Including data in tests

- Sometimes, you'll need to include toy datasets to use with your tests
- Let's write a new test for the `busiest_routes` function, to make sure it identifies the busiest routes
- We'll make a toy dataset to work with

## Including data in tests

- Put data files in `test/data/`
- Call it `passengers.csv`

```
from,to,Passengers,Distance
SFO,JFK,2,2586
RDU,CLT,3,130
JFK,SFO,2,2586
```

## Writing the test

- Write a test that reads the data in and tests that `busiest_routes` returns JFK->SFO as the busiest route
- Put the test in a new file
- You can get the path to your `passengers.csv` file with `os.path.join(os.path.dirname(__file__), "data", "passengers.csv")`

## Writing the test

```{python}
#| eval: false
import unittest
import pandas as pd
import os.path
import airlinestats

class TestBusiest(unittest.TestCase):
    def setUp (self):
        self.data = pd.read_csv(os.path.join(os.path.dirname(__file__), "data", "passengers.csv"))

    def test_busiest (self):
        routes = airlinestats.busiest_routes(self.data, "from", "to")
        routes = routes.sort_values("Passengers", ascending=False)
        self.assertEqual(routes.airport1.iloc[0], "JFK")
        self.assertEqual(routes.airport2.iloc[0], "SFO")
```

## When...

### to create a package

- When you are building reusable functions for a large project, or several projects
- When you want to employ automated testing

### not to create a package

- Simple analyses and one-off projects
- Projects with few interdependencies and where `stopifnot` testing is sufficient

## How to use your package

- Two ways to use your package: through `pip install` and through importing a local package
- `install` is the "right" way to use packages
- Importing a local package has advantages, though
    - Always in sync with your code - no installation step needed
    - Much scientific code is not very useful outside the project it was designed for

## Writing scripts using your package

- Scripts or notebooks can be in your package directory or elsewhere
- If you're installing your package, you can load it in your scripts with `import`
- Otherwise, the notebooks and python files need to be in the same place so that `import` knows where to find the package

## Where to put your scripts

- Single-project package - keep them in the same directory as the package
- Multiple-project package - have a separate directory/repository for each project using the package

## Environment management: `pip`, `conda`, and `poetry`

::: {.columns}
:::: {.column width="50%"}
- Pretty much every Python project relies on many external packages
    - `pandas`, `numpy`, etc.
- Code that works on your machine may fail on someone else's, if they have different package versions (or even different Python versions)
- The full set of package versions, along with the Python version, is an _environment_
- Most likely, you have just one environment on your system—the one `conda` shipped with
::::

:::: {.column width="50%"}
![&copy; [xkcd](https://xkcd.com/1987/)](xkcd_python_environment.png){fig-alt="XKCD comic showing a very convoluted directory structure with multiple Python i nstallations, with the caption 'My Python environment has become so degraded my laptop has been declared a Superfund site"}
::::
:::

## Environment management tools

- In scientific Python usage, `conda` is by far the most common environment management tool
- In other domains, `pip` and `poetry` are more common
- Here we'll talk about `conda`

## What is `conda`?

- Conda is a tool that manages the installation of Python packages and their dependencies
- It ensures version compatibility between installed packages
- It can record the exact versions of packages used, and reproduce that environment elsewhere

## What is a `conda` environment?

- A set of packages and a Python version separate from other `conda` environments
- It's good practice to have a separate `conda` environment for each project you're working on

## How do you create a `conda` environment?

- You can create one from the command line with the `conda create -n <name>` command
- But I prefer to specify my dependencies in an `environment.yml` file
- This way, I can document what packages I'm using, and restore them later

## What is an `environment.yml`

- The `environment.yml` file specifies the packages and versions needed for your project

```{yaml}
name: air-analysis
channels:
  - conda-forge
  - defaults
dependencies:
  - python=3.12.*
  - pandas>2.1,<3.0.0
  - ipython>8.17,<9.0.0
  - ipykernel>6.26,<7.0.0
```

## Creating `environment.yml`

- You can automatically generate an `environment.yml`, but I prefer to create one manually
- An autogenerated `environment.yml` file will pin _every_ package to a specific version, and prevent upgrades
    - It may also not work when changing operating systems or CPU types
    - Autogenerated files are useful for reproducibility; it's good practice to archive them when you finish your analysis so you can get back to exactly the package versions you had before
- I create my `environment.yml` with just the packages I'm actually using, and let `conda` figure out what other packages are needed to make them work

## Creating `environment.yml`

- Create an `environment.yml` file (using the "Text file" option in JupyterLab, or a text editor)
- The first line is the name of the environment

```{yaml}
name: air-analysis
```

## Creating `environment.yml`

- Next, we specify the _channels_ we want to use
- A channel is just a set of Conda packages
- If you just want to use the packages that came with Anaconda, you can leave this blank
- But if you want to use `conda-forge` (a community-driven, open-source packaging effort), we need to specify that

```{yaml}
channels:
    - conda-forge
    - defaults
```

## Creating `environment.yml`

- The main part of the file specifies our dependencies

```{yaml}
dependencies:
  - python=3.12.*
  - pandas~=2.1
  - ipython~=8.17
  - ipykernel~=6.26
```

- We want any patch release of Python 3.12
- We want a pandas version compatible with 2.1, but not pandas 3 or greater (there is not currently a pandas 3)
- etc.

## Semantic versioning

- Many projects use _semantic versioning_
- The version is `<major>.<minor>.<patch>`
- e.g. pandas 2.1.2 is major release 2, minor release 1, patch release 2
- New patch releases can only fix bugs
- New minor releases can fix bugs or add backwards-compatible features
- New major releases can make breaking changes
- Versions 0.*, all bets are off
    - But usually, breaking changes will be new minor versions

## Creating an environment from your `environment.yml`

- `conda env create -n environment-name -f environment.yml`
- Conda will "solve" the dependencies and install the requested packages
- If this is too slow, you can look into [mamba](https://mamba.readthedocs.io/en/latest/)

## Activating the environment

- When you activate an environment, running `python` etc. will refer to the packages installed in the environment
- Run `conda activate air-analysis`

## Using Jupyter with Conda environments

- There are two options for using JupyterLab with a conda environment
    - Install Jupyter inside your conda environment
    - Install `ipykernel` in your environment and create a Jupyter "kernel" you can access from another environment

## Using Jupyter with Conda environments

- Let's install `jupyterlab` into our conda environment
- Edit your `environment.yml` to include a dependency `jupyterlab~=4.0`
- Run `conda env update -f environment.yml -n air-analysis`
- Now you should be able to run `jupyter lab` and start up jupyter in your conda environment
    - You may want to shut down any currently running jupyter sessions
- This will also update outdated packages in your environment

## Exporting a conda environment

- Sometimes, you want to export the _exact_ set of packages you have installed, to ensure reproducibility or to move an environment to another machine
- To do this, run `conda env export -f environment-complete.yml`
- This will generate an `environment.yml` with _every_ package
- This will restrict any version upgrades, and may not work at all on other operating systems/CPU architectures
- But it will give you _exactly_ the packages you had installed when you exported it

## Documenting your code

- Functions in Python can have _docstrings_ that contain information about the function
- This is just a literal Python string that is the first line of the function definition
- Generally enclosed in triple-double-quotes (""" and """) so that it can extend to multiple lines
- It's a good practice to document every function, its arguments, assumptions it makes, etc.

## Documenting your code

- Let's add a docstring to our `busiest_routes` function
- It's typical to format your docstring onto multiple lines manually, rather than letting it wrap automatically

```{python}
def busiest_routes(dataframe, origcol="Origin", destcol="Dest"):
    """
    Calculate busiest routes from a dataframe of DB1B data. It expects disaggregate, ticket-level data with
    a Passengers column indicating how many passengers were on each ticket.

    Arguments:

    - dataframe -- required, the data to use
    - origcol -- optional, the column to use for representing the origin (default "Origin")
    - destcol -- optional, the column to use for representing the destination (default "Dest")
    """
    assert (dataframe.Passengers >= 1).all()
```

## Why not just write comments?

- You definitely should write comments throughout your functions to explain what they're doing
- But docstrings are special for two reasons
    - You can access a function's docstring after it's defined
        ```{python}
        print(busiest_routes.__doc__)
        ```

    - Documentation-generation software (e.g. MkDocs, Sphinx) can use the docstrings to generate function-level documentation
